# Split-Apply-Combine: I

The split-apply-combine paradigm can be concisely summarized using the diagram below (thanks hadley!)

<img src=http://inundata.org/R_talks/meetup/images/splitapply.png width=75%></img>

Base R has several functions that make this easy. Let us start by revisiting __Exercise 3__ from the previous lesson. This time, we will use the base function `aggregate` to carry out the computations. We use the formula interface provided by `aggregate`.

```{r}
bnames2_b = mutate(bnames2_b, tot = prop * births)
result <- aggregate(
  formula = tot ~ name, 
  data = bnames2_b, 
  FUN = sum
)
```

__Exercise 1__

What is the most popular `name` by `gender` between the years 2000 to 2008? (Hint: The `aggregate` function accepts a `subset` argument!) Once again, enter your guesses on Etherpad before starting out!

__Solution 1__

```{r}
result2 <- aggregate(
  formula = tot ~ name + sex,
  data = bnames2_b,
  FUN = sum,
  subset = (year >= 2000)
)
most_pop_boy <- arrange(subset(result2, sex == "boy"), desc(tot))[1,"name"]
most_pop_girl <- arrange(subset(result2, sex == "girl"), desc(tot))[1,"name"]
```

<a href="javascript:expandcollapse('solution1')">
   [+/-] Solution
</a><br>

<span class='posthidden' id='solution1'>
The most popular names between 2000 and 2008 are __`r most_pop_boy`__ and __`r most_pop_girl`__
</span>


So far, we have seen `split-apply-combine` applied in the context of data frames. However, you can think of similar problems for other data structures. Here are some examples.

1. Extract components by name from a `list`.
2. Compute the mean across rows/columns of a `matrix`.
3. Apply a `function` across multiple sets of arguments.

Base R has a family of functions, popularly referred to as the `apply` family to carry out such operations. 

__apply__

`apply` applies a function to each row or column of a matrix.

```{r, eval = TRUE}
m <- matrix(c(1:10, 11:20), nrow = 10, ncol = 2)
# 1 is the row index
# 2 is the column index
apply(m, 1, sum)
apply(m, 2, sum)
apply(m, 1, mean)
apply(m, 2, mean)
```

__lapply__


`lapply` applies a function to each element of a list


```{r}
my_list <- list(a = 1:10, b = 2:20)
lapply(my_list, mean)
```

__sapply__

`sapply` is a more user friendly version of `lapply` and will return a list of matrix where appropriate. Let us work with the same list we just created.

```{r, eval = TRUE}
my_list
x <- sapply(my_list, mean)
x
class(x)
```

__mapply__

Its more or less a multivariate version of `sapply`. It applies a function to all corresponding elements of each argument. 

```{r}
list_1 <- list(a = c(1:10), b = c(11:20))
list_1
list_2 <- list(c = c(21:30), d = c(31:40))
list_2
mapply(sum, list_1$a, list_1$b, list_2$c, list_2$d)
```


__tapply__

`tapply` applies a function to subsets of a vector.

```{r}
head(warpbreaks)
with(warpbreaks, tapply(breaks, list(wool, tension), mean))
```



__by__

`by` applies a function to subsets of a data frame.

```{r}
head(iris)
by(iris[, 1:2], iris[,"Species"], summary)
by(iris[, 1:2], iris[,"Species"], sum)
```



__replicate__

An extremely useful function to generate datasets for simulation purposes. The final arguments turns the result into a vector or matrix if possible.

```{r}
replicate(10, rnorm(10))
replicate(10, rnorm(10), simplify = TRUE)
```

While these functions in Base R get the job done, the inconsistent syntax often trips up users. Over the years, several alternatives have emerged, that aim to provide a simpler and more consistent syntax to operationalize `split-apply-combine`. In the next lesson, we will explore three packages in particular: `plyr`, `data.table` and `dplyr`.


<style>
  .posthidden {display:none} 
  .postshown {display:inline}
</style>

<script type="text/Javascript"> 

  function expandcollapse (postid) { 
    whichpost = document.getElementById(postid); 
   
     if (whichpost.className=="postshown") { 
      whichpost.className="posthidden"; 
     } 
     else { 
      whichpost.className="postshown"; 
     } 
  } 
</script>


