# Tidy Data

```{r setup, cache = F, echo = F}
library(knitr)
opts_chunk$set(tidy = F)
library(xtable)
options(xtable.type = 'html', xtable.include.rownames = FALSE)
```


__Question:__ Consider the following data below. How many variables does this dataset contain?

```{r tidy-data-1, echo = F}
dat <- data.frame(
  males = c(injured = 4, uninjured = 2), 
  females = c(injured = 1, uninjured = 5)
)
```

```{r tidy-data-2, echo = F, results = 'asis', comment = NA}

print(xtable(dat, digits = 0), html.table.attributes = "style='width:50%;'")
```

<a href="javascript:expandcollapse('solution1')">
   [+/-] Solution
 </a><br>

<span class='posthidden' id='solution1'>
The way the table is presented, it seems like there are only two variables. However, the correct answer is __3__. The variables are `injured?`, `count` and `gender`. 
</span>

> Raw data in the real-world is often `untidy` and poorly formatted. Furthermore, it may lack appropriate details of the study. Correcting data in place can be a dangerous exercise since the original raw data would get overwritten and there would be no way to audit this process or recover from mistakes made during this time. A good data practice would be to maintain the original data, but use a programmatic script to clean it, fix mistakes and save that cleaned dataset for further analysis. 


### What is Tidy Data?

A dataset is said to be tidy if it satisfies the following conditions

1. observations are in rows
2. variables are in columns
3. contained in a single dataset.

Tidy data makes it easy to carry out data analysis.

### Explore Messy Data

Let us explore some common causes of messiness by inspecting a few datasets. 

__Income Distribution by Religion__


Our first dataset is based on a survey done by Pew Research that examines the relationship between [income and religious affiliation](http://www.pewforum.org/2009/01/30/income-distribution-within-us-religious-groups/). 

Read the dataset into your R session and inspect the first few rows to assess if it is tidy.

```{r tidy-data-pew}
pew <- read.delim(
  file = "http://stat405.had.co.nz/data/pew.txt",
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = F
)
```

```{r echo = F, results = 'asis', comment = NA}
xtable(pew[1:4,1:10], digits = 0)
```

<a href="javascript:expandcollapse('solution1')">
   [+/-] Solution
 </a><br>

<span class='posthidden' id='solution1'>
Except for `religion`, the rest of the columns headers are actually values of a lurking variable `income`. This dataset violates the second rule of tidy data.
</span>

__Tuberculosis Incidence__

```{r tidy-data-tb}
tb <- read.csv(
  file = "http://stat405.had.co.nz/data/tb.csv",
  header = TRUE, 
  stringsAsFactors = FALSE
)
```


```{r echo = F, results = 'asis', comment = NA}
print(xtable(tail(tb[,1:7], 5), digits = 0), 
  html.table.attributes='style="width:90%;"'
)
```

<a href="javascript:expandcollapse('solution2')">
   [+/-] Solution
 </a><br>

<span class='posthidden' id='solution2'>
Except for `iso2` and `year`, the rest of the columns headers are actually values of a lurking variable, in fact combination of __two__ lurking variables, `gender` and `age`.
</span>

__Weather Data__

```{r tidy-data-weather}
weather <- read.delim(
 file = "http://stat405.had.co.nz/data/weather.txt",
 stringsAsFactors = FALSE
)
```

```{r echo = F, results = 'asis', comment = NA}
xtable(tail(weather[,1:11], 5), digits = 0)  
```


<a href="javascript:expandcollapse('solution3')">
   [+/-] Solution
</a><br>

<span class='posthidden' id='solution3'>
This dataset seems to have two problems. First, it has variables in the rows in the column `element`. Second, it has a variable `d` in the column header spread across multiple columns.
</span>






### Causes of Messiness

- Column headers are values, not variable names 
- Multiple variables are stored in one column 
- Variables are stored in both rows and columns 
- Multiple types of experimental unit stored in the same table 
- One type of experimental unit stored in  multiple tables



<style>
  .posthidden {display:none} 
  .postshown {display:inline}
</style>

<script type="text/Javascript"> 

  function expandcollapse (postid) { 
    whichpost = document.getElementById(postid); 
   
     if (whichpost.className=="postshown") { 
      whichpost.className="posthidden"; 
     } 
     else { 
      whichpost.className="postshown"; 
     } 
  } 
</script>
